---
title: "Methods for JS Lewis analysis"
author: "Brandon"
date: "2025-02-20"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Methods

This is an individual based model describing the estimate of carcass abundance in the Lewis River. This model is different than that of Schwarz. The goal is to estimate the total carcass abundance (i.e., $\lambda$ \ref{GenericLikelihood})


To understand the likelihood, it is best to start with a model that has no spatial component.
\begin{equation} \label{eq:Likelihood}
  L = P(B_{total}|\lambda p)\prod_i^{B_{total}} P(t_i|\boldsymbol{\pi})P(D_{i,t}|t_i,\phi_i,p_i)P(T_{i,t}|D_{i,t},\psi_i)P(R_{i,t\prime}|T_{i,t},\phi_i,p_i) 
\end{equation}

Breaking down the likelihood into it's components, the probability of the observed total number of carcasses is
\begin{equation}\label{P_PopulationSize}
  P(B_{total}|\lambda p)
\end{equation}
where,  $\lambda$ is total population size, and p is the capture probability of a carcass.
The probability of the arrival process is,
\begin{equation}\label{P_arrival}
  P(t_i|\boldsymbol{\pi})
\end{equation}
where, $t_i$ is a latent effect, and $\boldsymbol{\pi}$ vector of arrival probabilities for each time-step.
The probability of the initial detection for any carcass, 
\begin{equation}\label{P_initial detection}
  P(D_{i,t}|t_i,\phi_i,p_i)
\end{equation}
where, $D_{i,t} = 1$, $t_i$ is the latent estimate of arrival time, $phi_i$ is the survival probability, and $p_i$ is the detection probability.
The probability of carcass being tagged is,
\begin{equation}\label{P_tagging}
  P(T_{i,t}|D_{i,t},\psi)
\end{equation}
where, $T_{i,t}=1$ is a carcass was tagged, $\psi$ is the tagging rate.
Finally, there is probability of recapture,

\begin{equation}\label{P_recapture}
  P(R_{i,t\prime}|T_{i,t},\phi_i,p_i)
\end{equation}

where, $R_{i,t\prime} = 1$ for a carcass recaptured on day $t\prime$.

## Create the data

We begin by reading the data and transforming some of the output to make it easier to work in RTMB.

```{r, warning=FALSE}
#Data list
library(dplyr)
library(tidyr)

d <- read.csv("data/simpleData2.csv") %>%
  mutate(t_wk = lubridate::week(lubridate::mdy(TagDate)),
         r_wk = lubridate::week(lubridate::mdy(RecapDate)),
         t_yr = lubridate::year(lubridate::mdy(TagDate)),
         r_yr = lubridate::year(lubridate::mdy(RecapDate))) %>%
  filter(t_yr == 2024) %>% 
  filter(t_wk > 10) %>%
  mutate(t_k = t_wk - min(t_wk) + 1,
         r_k = r_wk - min(t_wk) + 1) %>%
  mutate(t_l = TagState,
         r_l = RecapState) %>%
  filter(is.na(r_wk) | r_k>0) %>% 
  mutate(tag = ifelse(Tag1=="",FALSE,TRUE)) %>%
  group_by(t_k,r_k,t_l,r_l,tag) %>%
  summarise(n = n())
```


## Just model the CJS $P(R_{i,t\prime}|T_{i,t},\phi_i,p_i)$

It is possible to breaks down the analysis in to separate components. We can start by just looking at the Cormack-Jolly-Seber part of the model. For those tagged individuals that are tagged, we can estimate survival and detection probability. For now, we can simply assume all fish are equal and the detection and survival is constant across time and location. The probability of 

I am going to use a matrix-algebra approach to analyzing the data. And start with the simplest data set possible.

```{r}
data <- list(t_l = d$t_l, #tagging location
             r_l = d$r_l, #recapture location
             t_k = d$t_k, #tagging week
             r_k = d$r_k,
             tag = d$tag,
             n = d$n) #recapture week, last week if not recapture

# Initial parameter values
parameters <- list(
  phi_par = 0,
  p_par = 0
  )
```

Now lets create the likelihoods function given the data.
```{r, echo = FALSE, warning=FALSE, message = FALSE}
library(RTMB)
```

```{r, message=FALSE, warning=FALSE}
f <- function(parms){

  RTMB::getAll(data,
               parms)
  
  #Negative
  nll <- rep(0,length(d$t_l))
  
  #Survival
  phi <- matrix(0,2,2)
  phi[1,1] <- -exp(phi_par)
  phi[1,2] <- -phi[1,1]
  phi <- Matrix::expm(phi)
  
  #Detection probability
  p <- matrix(0,2,2)
  p[1,1] <- -exp(p_par)
  p[1,2] <- -p[1,1]
  p <- Matrix::expm(p)
  
  #This is the amount of time to first detection
  for(i in 1:length(t_l)){
    
    #Accumulator
    m <- matrix(0,2,2)
    diag(m) <- 1
    
    #Initial state
    delta <- rep(0,2)
    delta[1] <- 1
    if(tag[i] == TRUE ){
      if(is.na(r_k[i])){
        t_prime <- 16
      }else{
        t_prime <- r_k[i]
      }
      for(j in (t_k[i]+1):t_prime){
        if(j<t_prime){ #not detected
          m <- m %*% phi %*% diag(p[,2])
        }else{#last observation
          if(is.na(r_k[i])){#not detected
            m <- m %*% phi %*% diag(p[,2])
          }else{#Detected
            m <- m %*% phi %*% diag(p[,1])
          }
        }
      }
      tmp_phi <- phi * 0
      tmp_phi[,2] <- phi[,2]
      m <- m %*% tmp_phi

    }
    nll[i] <- log(t(delta) %*% m %*% rep(1,2))
  }

  RTMB::REPORT(nll)  
  RTMB::REPORT(phi)  
  RTMB::REPORT(p)  
  return(-sum(nll*n))
  # return(0)
}

```

Next we can optimize.
```{r, message=FALSE, warning=FALSE, echo = FALSE}
obj <- RTMB::MakeADFun(f,
                       parameters,
                       silent = TRUE)
opt <- nlminb(obj$par,
              obj$fn,
              obj$gr)
rep <- obj$report()

```

### Results

Finally, we can look at the results of $\phi$ and $p$ matrixes,

```{r}
#Survival/persistence
print(round(rep$phi,2))

#Survival/persistence
print(round(rep$p,2))

```

In this simple model, for the tagged individuals, the "persistence/survival" between time steps is `r round(rep$phi[1,1],2)` and the detection probability is `r round(rep$p[1,1],2)`.


## Just model the CJS $P(R_{i,t\prime}|T_{i,t},\phi_i,p_i)$ plus tagging rate $P(T_{i,t}|D_{i,t},\psi)$ 

```{r}
# Initial parameter values
parameters <- list(
  phi_par = 0,
  p_par = 0,
  psi_par = 0
  )
```

```{r, message=FALSE, warning=FALSE}
f <- function(parms){

  RTMB::getAll(data,
               parms)
  
  #Negative
  nll <- rep(0,length(d$t_l))
  nll_tag <- rep(0,length(d$t_l))   
  
  #Survival
  phi <- matrix(0,2,2)
  phi[1,1] <- -exp(phi_par)
  phi[1,2] <- -phi[1,1]
  phi <- Matrix::expm(phi)
  
  #Detection probability
  p <- matrix(0,2,2)
  p[1,1] <- -exp(p_par)
  p[1,2] <- -p[1,1]
  p <- Matrix::expm(p)

  #This is the amount of time to first detection
  for(i in 1:length(t_l)){
  
    #Probability of tagging
    nll_tag[i] <-  RTMB::dbinom(tag[i]*n[i],
                             n[i],
                             RTMB::plogis(psi_par), 
                             log = TRUE) 
    #Accumulator
    m <- matrix(0,2,2)
    diag(m) <- 1
    
    #Initial state
    delta <- rep(0,2)
    delta[1] <- 1
    if(tag[i] == TRUE ){
      if(is.na(r_k[i])){
        t_prime <- 16
      }else{
        t_prime <- r_k[i]
      }
      for(j in (t_k[i]+1):t_prime){
        if(j<t_prime){ #not detected
          m <- m %*% phi %*% diag(p[,2])
        }else{#last observation
          if(is.na(r_k[i])){#not detected
            m <- m %*% phi %*% diag(p[,2])
          }else{#Detected
            m <- m %*% phi %*% diag(p[,1])
          }
        }
      }
      tmp_phi <- phi * 0
      tmp_phi[,2] <- phi[,2]
      m <- m %*% tmp_phi

    }
    nll[i] <- log(t(delta) %*% m %*% rep(1,2))
  }

  psi <- RTMB::plogis(psi_par)
  
  RTMB::REPORT(nll)  
  RTMB::REPORT(nll_tag)
  RTMB::REPORT(phi)  
  RTMB::REPORT(p)  
  RTMB::REPORT(psi)
  return(-sum(nll*n) - sum(nll_tag))
  # return(0)
}

```

Next we can optimize.

```{r, message=FALSE, warning=FALSE, echo = FALSE}
obj <- RTMB::MakeADFun(f,
                       parameters,
                       silent = TRUE)
opt <- nlminb(obj$par,
              obj$fn,
              obj$gr)
rep <- obj$report()
```

### Results

Not surprisingly, the values for the survival ($\phi$) and detection probability ($p$), `r round(rep$phi,2)` and `r round(rep$p,2)`, respectively, remain unchanged, while the estimate of tagging rate ($\psi$) is equal to `r round(rep$psi,2)`. The estimate for $\psi$ is identical to `r sum(data$tag*data$n)`, the number of carcasses tagged, divided by the `r sum(data$n)`, the number of carcasses observed. 


## Just model the CJS $P(R_{i,t\prime}|T_{i,t},\phi_i,p_i)$, plus tagging rate $P(T_{i,t}|D_{i,t},\psi)$, plus random effects for arrival timing.

```{r}
# Initial parameter values
parameters <- list(
  phi_par = 0,
  p_par = 0,
  psi_par = 0,
  par_alpha = 0,
  par_beta = 0,
  re_ti = rep(0, length(data$t_l)),
  par_lambda = 10
  )
```

```{r, message=FALSE, warning=FALSE}
f <- function(parms){

  RTMB::getAll(data,
               parms)
  
  #Negative
  nll <- rep(0,length(d$t_l))
  nll_tag <- rep(0,length(d$t_l))   
  nll_time <- rep(0, length(d$t_l))
  
  #Survival
  phi <- matrix(0,2,2)
  phi[1,1] <- -exp(phi_par)
  phi[1,2] <- -phi[1,1]
  phi <- Matrix::expm(phi)
  
  #Detection probability
  p <- matrix(0,2,2)
  p[1,1] <- -exp(p_par)
  p[1,2] <- -p[1,1]
  p <- Matrix::expm(p)

  #Likleihood of total carcasses
  nll_pop <- RTMB::dpois(sum(n), exp(par_lambda) * p[1,1])
  
  #This is the amount of time to first detection
  for(i in 1:length(t_l)){
  
    #Probability of tagging
    nll_tag[i] <-  RTMB::dbinom(tag[i]*n[i],
                             n[i],
                             RTMB::plogis(psi_par), 
                             log = TRUE) 

    #Arrival timing    
    nll_time[i] <-  RTMB::dnorm(exp(re_ti[i]),
                         exp(par_alpha),
                         exp(par_beta), 
                         log = TRUE) 

    #Accumulator
    m <- matrix(0,2,2)
    diag(m) <- 1
    
    #Initial state
    delta <- rep(0,2)
    delta[1] <- 1

    #Likelihood of first observation, based on predicted number of time steps the carcass has been in the river
    m <- m %*% Matrix::expm(phi * exp(re_ti[i])) %*% diag(p[,1])
    
    #All subseqeunt observations for tagged individuals
    if(tag[i] == TRUE ){
      
      #t_prime is the last capture period
      if(is.na(r_k[i])){
        t_prime <- 16
      }else{
        t_prime <- r_k[i]
      }
      
      for(j in (t_k[i]+1):t_prime){ #loop over time steps
        if(j<t_prime){ #not detected
          m <- m %*% phi %*% diag(p[,2])
        }else{#last observation
          if(is.na(r_k[i])){#not detected on last observation
            m <- m %*% phi %*% diag(p[,2])
          }else{#Detected  on last observation
            m <- m %*% phi %*% diag(p[,1])
          }
        }
      }
      
      # tmp_phi <- phi * 0
      # tmp_phi[,2] <- phi[,2]
      # m <- m %*% tmp_phi

    }
    nll[i] <- log(t(delta) %*% m %*% rep(1,2))
  }

  psi <- RTMB::plogis(psi_par)
  beta <- exp(par_beta)
  alpha <- exp(par_alpha)
  lambda <- exp(par_lambda)
  
  RTMB::REPORT(nll)  
  RTMB::REPORT(nll_tag)
  RTMB::REPORT(nll_time)
  RTMB::REPORT(phi)  
  RTMB::REPORT(p)  
  RTMB::REPORT(psi)
  RTMB::REPORT(alpha)
  RTMB::REPORT(beta)
  RTMB::REPORT(lambda)
  return(-sum(nll*n) - sum(nll_tag) - sum(nll_time) - nll_pop)
  # return(0)
}

```

Next we can optimize.

```{r, message=FALSE, warning=FALSE, echo = FALSE}
obj <- RTMB::MakeADFun(f,
                       parameters,
                       random = c("re_ti"),
                       # map = list(re_ti = as.factor(rep(NA, length(parameters$re_ti))),
                       #            par_alpha = as.factor(NA),
                       #            par_beta = as.factor(NA)),
                       silent = FALSE)
opt <- nlminb(obj$par,
              obj$fn,
              obj$gr)
rep <- obj$report()
```

### Results

Not surprisingly, the values for the survival ($\phi$) and detection probability ($p$), `r round(rep$phi,2)` and `r round(rep$p,2)`, respectively, remain unchanged, while the estimate of tagging rate ($\psi$) is equal to `r round(rep$psi,2)`. The estimate for $\psi$ is identical to `r sum(data$tag*data$n)`, the number of carcasses tagged, divided by the `r sum(data$n)`, the number of carcasses observed. 

# Tables

```{r kable-table, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)

param_table <- data.frame(
  Symbol = c("$\\lambda$", 
             "$\\boldsymbol{\\pi}$", 
             "$p$", 
             "$\\phi$", 
             "$\\psi$", 
             "$t_i$", 
             "$B_{\\text{total}}$", 
             "$T_{i,t}$", 
             "$D_{i,t}$", 
             "$R_{i,t'}$"),
  Description = c("True total number of carcasses", 
                  "Arrival time probabilities", 
                  "Detection probability", 
                  "Persistence probability", 
                  "Tagging probability", 
                  "Carcass arrival time",
                  "Observed total number of carcasses", 
                  "Tagging indicator (1 if tagged, 0 otherwise)", 
                  "Detection indicator (1 if detected, 0 otherwise)", 
                  "Recapture indicator (1 if recaptured, 0 otherwise)"),
  Type = c("Parameter (to be estimated)", 
           "Parameter (to be estimated)", 
           "Parameter (to be estimated)", 
           "Parameter (to be estimated)", 
           "Parameter (to be estimated)", 
           "Latent (to be estimated)", 
           "Data (observed)", 
           "Data (observed)", 
           "Data (observed weekly)", 
           "Data (observed)")
)

```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#Read in data
library(tidyr)
library(dplyr)
library(kableExtra)

d <- read.csv("data/simpleData2.csv") %>%
  mutate(t_wk = lubridate::week(lubridate::mdy(TagDate)),
         r_wk = lubridate::week(lubridate::mdy(RecapDate))) %>%
  filter(t_wk > 10) %>%
  mutate(t_k = t_wk - min(t_wk) + 1,
         r_k = r_wk - min(t_wk) + 1) %>%
  mutate(t_l = TagState,
         r_l = RecapState) %>%
  mutate(tag = ifelse(Tag1=="",FALSE,TRUE)) %>%
  group_by(t_k,r_k,t_l,r_l,tag) %>%
  summarise(n = n())

```


```{r, echo = FALSE, warning=FALSE, message=FALSE}

# kable(param_table, escape = FALSE, caption = "Table 1: Model Parameters and Data") %>%
#   kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
# 
# kable(head(d), caption = "Table 2. Fish survey data from the Lewis River from 2024 includes the tag week (t_k), recapture week (r_k), tag location (t_l), recapture location (r_l), whether the fish was tagged (tag), and the number of carcasses (n), with that particular unique spatiotemporal combination of tagging and recapture.") %>%
#   kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

```
